---
title: "Final analysis (1/2) — Inspect count matrices, gene names, metadata, Effector CD4+ subset"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.path = "results/figure-html-01/",
  dpi = 120
)
dir.create("results/figure-html-01/", recursive = TRUE, showWarnings = FALSE)
dir.create("exported", showWarnings = FALSE)

## Standalone repo: `data/` here. Nested in larger project: `../data/`.
repo_root <- if (dir.exists("data")) normalizePath(".") else normalizePath("..")
if (!dir.exists(file.path(repo_root, "data"))) {
  stop("Place GEO count matrices in data/ (see data/README.md).")
}

if (!rmarkdown::pandoc_available()) {
  for (brew_bin in c("/opt/homebrew/bin", "/usr/local/bin")) {
    if (file.exists(file.path(brew_bin, "pandoc"))) {
      Sys.setenv(PATH = paste(brew_bin, Sys.getenv("PATH"), sep = .Platform$path.sep))
      break
    }
  }
}

## Visible in Console + HTML while knitting (so long steps are not “silent”).
.run_step <- function(label) {
  cat("\n[", format(Sys.time(), "%H:%M:%S"), "] ", label, "\n", sep = "")
  flush.console()
}
```

This notebook loads **ATAC** and **RNA raw counts**, renames RNA columns **GSM → `sample_id`**, rebuilds **`final_analysis/sample_metadata.txt`** from the supplementary QC sheet, and exports **Effector CD4+** sample IDs for **`02_effector_cd4_limma_PCA_integration.Rmd`**. **Gene-level annotation (Entrez → SYMBOL / GENETYPE / ENSEMBL) is deferred to notebook 02** and applied only to genes that pass the RNA **CPM** filter, matching the earlier atlas notebooks’ strategy (no whole-matrix `mapIds` on ~40k IDs).

**Runtime:** this notebook should finish in **under a minute** (peek-only matrix I/O, no full ATAC/RNA decompression for row counts). **Where to watch progress:** in RStudio use **Knit** and watch the **R Markdown** jobs pane; from the shell use `rmarkdown::render("01_matrices_metadata.Rmd", quiet = FALSE)` so knitr prints **`i/n [chunk-name]`** as each chunk runs. Chunks also call **`.run_step("…")`** (timestamps appear in the rendered HTML code output).

```{r libraries}
library(readxl)
library(dplyr)
library(knitr)
library(org.Hs.eg.db) ## only for small preview chunk below
```

# Paths

```{r paths}
.run_step("01 — paths and file checks")
path_atac <- file.path(repo_root, "data", "GSE118189_ATAC_counts.txt.gz")
path_rna_gz <- file.path(repo_root, "data", "GSE118165_raw_counts_GRCh38.p13_NCBI.tsv.gz")
path_rna_plain <- file.path(repo_root, "data", "GSE118165_raw_counts_GRCh38.p13_NCBI.tsv")
path_gsm_map <- file.path(repo_root, "data", "GSE118165_GSM_to_sample_id.tsv")
path_xlsx <- file.path(repo_root, "data", "Supplementary_tables.xlsx")
qc_sheet <- "ATAC-seq samle QC"
path_meta_out <- if (repo_root == normalizePath(".")) "sample_metadata.txt" else file.path(repo_root, "final_analysis", "sample_metadata.txt")
path_cd4_samples <- file.path(if (repo_root == normalizePath(".")) "exported" else file.path(repo_root, "final_analysis", "exported"), "effector_cd4_sample_ids.txt")

path_rna <- if (file.exists(path_rna_gz)) path_rna_gz else path_rna_plain
rna_open <- if (grepl("\\.gz$", path_rna)) function() gzfile(path_rna, "rt") else function() file(path_rna, "rt")

stopifnot(
  file.exists(path_atac),
  file.exists(path_rna),
  file.exists(path_gsm_map),
  file.exists(path_xlsx)
)
```

# ATAC matrix — structure

```{r atac-inspect}
.run_step("01 — ATAC: peek (header + 8 rows only)")
atac_header <- read.table(
  gzfile(path_atac),
  header = TRUE,
  sep = "\t",
  row.names = 1,
  check.names = FALSE,
  nrows = 1L
)
atac_samples <- colnames(atac_header)

atac_peek <- read.table(
  gzfile(path_atac),
  header = TRUE,
  sep = "\t",
  row.names = 1,
  check.names = FALSE,
  nrows = 8L
)

## Row counts: omit full-file line counts (gunzip | wc -l on multi‑GB tables can take many minutes).
n_atac_rows <- NA_integer_

kable(data.frame(
  file = basename(path_atac),
  n_peaks = n_atac_rows,
  n_samples = length(atac_samples)
), caption = "ATAC count matrix (n_peaks omitted in this notebook to avoid decompressing the entire file; see `dim(atac_full)` in notebook 02 after load)")

cat("\nExample peak IDs (row names):\n")
print(head(rownames(atac_peek), 5))

cat("\nExample sample columns:\n")
print(head(atac_samples, 10))

cat("\nCorner: 6 peaks × 6 samples\n")
print(atac_peek[, head(colnames(atac_peek), 6), drop = FALSE])
```

# RNA raw counts — GSM → `sample_id`, structure

```{r rna-load-rename}
.run_step("01 — RNA: peek (header + 12 rows), GSM → sample_id")
rna_header <- read.table(
  rna_open(),
  header = TRUE,
  sep = "\t",
  row.names = 1,
  check.names = FALSE,
  nrows = 1L
)
gsm_map <- read.table(path_gsm_map, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
stopifnot(all(colnames(rna_header) %in% gsm_map$GSM))

rna_peek <- read.table(
  rna_open(),
  header = TRUE,
  sep = "\t",
  row.names = 1,
  check.names = FALSE,
  nrows = 12L
)
colnames(rna_peek) <- gsm_map$sample_id[match(colnames(rna_peek), gsm_map$GSM)]

n_rna_rows <- NA_integer_

rna_samples <- gsm_map$sample_id[match(colnames(rna_header), gsm_map$GSM)]

kable(data.frame(
  file = basename(path_rna),
  n_genes = n_rna_rows,
  n_samples = length(rna_samples),
  row_id_type = "NCBI Entrez GeneID"
), caption = "RNA raw count matrix (n_genes omitted here to avoid streaming the whole file; notebook 02 reports `nrow(rna_full)` after load)")

cat("\nExample Entrez row names:\n")
print(head(rownames(rna_peek), 8))

cat("\nExample samples:\n")
print(head(colnames(rna_peek), 8))

cat("\nCorner: 8 genes × 6 samples (integer counts)\n")
print(rna_peek[, head(colnames(rna_peek), 6), drop = FALSE])

cat("\nValue summary (peek matrix):\n")
print(summary(as.vector(as.matrix(rna_peek))))
```

# Gene IDs — preview only (full annotation in notebook 02 after CPM filter)

```{r gene-annotation-preview}
.run_step("01 — mapIds preview (~40 Entrez only)")
## Small `mapIds` call on the peek matrix row names only (not the whole genome).
prev_ids <- head(unique(rownames(rna_peek)), 40L)
prev_sym <- mapIds(org.Hs.eg.db, prev_ids, "SYMBOL", "ENTREZID", multiVals = "first")
kable(
  data.frame(ENTREZID = prev_ids, SYMBOL = unname(prev_sym), stringsAsFactors = FALSE),
  caption = "Example Entrez → SYMBOL (notebook 02 annotates only CPM-passing genes)"
)
```

# Overlap ATAC vs RNA column names

```{r overlap-cols}
.run_step("01 — sample ID overlap ATAC vs RNA")
in_both <- intersect(atac_samples, rna_samples)
kable(data.frame(
  n_atac_only = length(setdiff(atac_samples, rna_samples)),
  n_rna_only = length(setdiff(rna_samples, atac_samples)),
  n_both = length(in_both)
))
```

# Supplementary QC → `sample_metadata.txt`

```{r build-metadata}
.run_step("01 — build sample_metadata.txt from supplementary QC")
qc <- read_excel(path_xlsx, sheet = qc_sheet)

qc_use <- qc %>%
  mutate(
    condition = case_when(
      .data$condition == "U" ~ "resting",
      .data$condition == "S" ~ "stimulated",
      TRUE ~ NA_character_
    ),
    in_atac = .data$sample %in% atac_samples,
    in_rna = .data$sample %in% rna_samples
  )

bad_cond <- qc_use %>% filter(is.na(.data$condition))
if (nrow(bad_cond) > 0) {
  print(unique(qc$condition))
  stop("Unexpected condition codes (expected U / S).")
}

meta_out <- qc_use %>%
  transmute(
    sample_id = .data$sample,
    cell_type = .data$cell_type,
    condition = .data$condition,
    donor = .data$donor,
    in_atac = .data$in_atac,
    in_rna = .data$in_rna
  ) %>%
  arrange(.data$cell_type, .data$donor, .data$condition)

write.table(
  meta_out,
  file = path_meta_out,
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)
cat("Wrote", nrow(meta_out), "rows →", path_meta_out, "\n")

meta_check <- read.table(
  path_meta_out,
  header = TRUE,
  sep = "\t",
  row.names = 1,
  check.names = FALSE
)
stopifnot(all(meta_check$in_atac == (rownames(meta_check) %in% atac_samples)))
stopifnot(all(meta_check$in_rna == (rownames(meta_check) %in% rna_samples)))

kable(head(meta_check, 12), caption = "sample_metadata.txt (read-back)")
```

# Effector CD4+ samples for notebook 02

```{r effector-cd4-export}
.run_step("01 — export Effector CD4+ sample list for notebook 02")
cd4_all <- rownames(meta_check)[meta_check$cell_type == "Effector_CD4pos_T"]
cd4_joint <- intersect(
  intersect(cd4_all, atac_samples),
  rna_samples
)

writeLines(cd4_joint, path_cd4_samples)
cat("Effector CD4+ samples (in both matrices):", length(cd4_joint), "\n")
kable(as.data.frame.matrix(table(
  meta_check[cd4_joint, "donor", drop = TRUE],
  meta_check[cd4_joint, "condition", drop = TRUE]
)), caption = "Donor × condition (Effector CD4+, both assays)")

cat("\nWrote:", path_cd4_samples, "\n")
```

# Session info

```{r session-info}
.run_step("01 — sessionInfo()")
sessionInfo()
```
